Online methods

### 2013-09-12 docs/manuscript/methods.tex
work continues

\subsection{Simulations}

@@

For each value of combination of the parameters $p_{1}, q_{1}, r, n$ 1000 simulations were performed and the sampling mean and sampling standard deviation of $\hat{r}, \hat{r}^2, \hat{r}^4, \hat{r}^6, \hat{r}^8$ were recorded.

@@

These simulations show that sampling variance of $r^x$ increases as $x$ increases. Detection of epistatis is highly dependent upon high $r^2$. Amongst the discovery SNPs there will be a mixture of interactions where observed SNPs are either in true high LD with causal variants, or will have highly inflated sample $\hat{r}^{x}$ compared to the population $r^x$. Supposing that SNPs

>>>>

\subsection{Simulation 1}

@@

For each value of combination of the parameters $p_{1}, q_{1}, r, n$ 1000 simulations were performed and the sampling mean and sampling standard deviation of $\hat{r}, \hat{r}^2, \hat{r}^4, \hat{r}^6, \hat{r}^8$ were recorded. It was observed that sampling variance increases for increasing $x$ in $\hat{r}^x$.

\subsection{Simulation 2}

We assume that the discovery SNP pairs are ascertained (from a very large number of tests) have high $\hat{r}$ between observed SNPs and causal variants because otherwise power of detection would be low. We can hypothesis that the distribution of $\hat{r}$ in this ascertained sample will be a mixture of $r$ that is high and $r$ that is lower but with ascertained higher values from sampling. Therefore, we would expect those with truly high $r$ to have a higher replication rate in independent datasets, and those with ascertained high $\hat{r}$ to have lower replication because resampling is unlikely to result in the same extreme ascertainment. To obtain empirical estimates of $\hat{r}$ in discovery and replication datasets we conducted the following simulation.

\begin{enumerate}
\item Using 1000 genomes data (phase 1, version 3, 379 European samples) we selected the 528,509 ``markers'' used in the original discovery analysis, plus 100,000 randomly chosen ``causal variants'' (CVs) with minor allele frequence $> 0.05$.
\item The 379 individuals were split into discovery (190) and replication (189) sets.
\item For each CV the marker with the maximum $\hat{r}^2_{D}$ from the marker panel was recorded in the discovery set. This marker was known as the ``discovery marker'' (DM).
\item The $\hat{r}^2_{R}$ for each CV/DM pair was then calculated in the replication set where the discovery LD was ascertained to be high, such that $\hat{r}^2_{D} > 0.9$.
\end{enumerate}

We observed that there was an average decrease in $\hat{r}^x_{R}$ relative to $\hat{r}^x_{D}$, and that this decrease was larger with increasing $x$. We observed that $(\hat{r}^2_{R} - \hat{r}^2_{D}) / \hat{r}^2_{D} = 0.029$ whereas $(\hat{r}^8_{R} - \hat{r}^8_{D}) / \hat{r}^8_{D} = 0.092$. The average drop in in replication $\hat{r}^8$ was 3 times higher than the drop in $\hat{r}^2$.

@@

Simulation 1 shows that sampling variance of $r^x$ increases as $x$ increases. Detection of epistatis is highly dependent upon high $\hat{r}$. Amongst the discovery SNPs there will be a mixture of interactions where observed SNPs are either in true high LD with causal variants, or will have highly inflated sample $\hat{r}^{x}$ compared to the population $r^x$. Simulation 2 shows that as $x$ gets larger, the average decrease in $\hat{r}^x$ between discovery and replication becomes larger, likely to be a result of ascertained high $\hat{r}$ in the discovery and increased sampling variance with increasing $x$ in the replication. These results demonstrate that assuming all else is equal, sampling variance of $r$ alone will reduce the replication rate of epistatic effects compared to additive effects.
